\myTitle

### 2024-07-05 tex/body-text-of-paper.tex
Edits in the Introduction

PBT has been remarkably successful in fisheries
management, having been used for over a decade in the management of Chinook salmon
and steelhead trout in major basins of the Columbia River \cite{steele2019parentage,horn2023multigeneration}, and having been employed to dramatically further our
understanding of the genetic inheritance of key traits in salmonids
\citep{abadia2013large,beulke2023distinct}. However, PBT is just one subset of a whole
family of statistical genetic methods employing relationship inference to learn about
populations.  For example inference of the full siblings amongst a sample of fish
provides information about the effective number of adults producing offspring
\citep{waples2011inbreeding,wang2023estimating}, which can
be a valuable source of information when sampling of the adults is not possible.

Here, we present a reference baseline for Chinook salmon, targeted to the complex population 
structure within the Central Valley of California. Chinook salmon of the two main river basins---the 
Sacramento and the San Joaquin---within the Central Valley exhibit the greatest run-timing diversity
within the species.  With four recognized ecotypes, delineated primarily on the basis of run timing
(fall-, late-fall-, winter-, and spring-run), adult Chinook salmon can be found migrating or residing
in freshwater every month of the year in California. \comment{NEED MORE HERE. Certainly we should cite
\citet{yoshiyama1998historical}.}

Chinook salmon are the largest of the Pacific salmonids and have historically been the target of extremely high-value and culturally important fisheries. Moreover, because of their high degree of ecotypic variation, they have provided fishery opportunities during many different seasons and geographic locations. However, recent declines in population sizes, from the Yukon River in Arctic north, to the southern extent of their range in California has led to almost complete fishery closures  throughout most of the northeastern Pacific to protect less productive stocks. The co-occurence of fish from relatively productive and relict populations reaches its paragon in California, with the largest remaining ocean fisheries for Chinook salmon targeting the Central Valley fall-run stock that spawns in different tributaries of the same river basin as the highly endangered and phenotypically distinct winter-run stock and the threatened spring-run stocks. In spite of their distinct phenotypes and life history patterns, the four ecotypes of Chinook salmon in the Central Valley are very closely related \cite{clemento2014evaluation} and share recent common ancestry that is independent of ongoing migration between these lineages. As such, GSI has been particularly challenging in the CCV, with at least one of the distinct ecotypes, late-fall-run, unresolvable with previous GSI marker sets \cite{seeb2007development,clemento2014evaluation} and with unsatisfactory power for discriminating the protected spring-run stocks from the harvested fall-run stock.

In the following we present a new reference baseline for both GSI and PBT in California.
We describe how we identified and compiled the markers, including several new markers
that are particularly effective at distinguishing some of the closely related groups,
such as late-fall and fall-run fish.  We also provide a comprehensive analysis of this
set of markers for inference of parents and siblings within different groups of populations.

>>>>

PBT has been remarkably successful in fisheries
management, having been used for over a decade in the management of Chinook salmon
and steelhead trout in major basins of the Columbia River \citep{steele2019parentage,horn2023multigeneration}, and having been employed to dramatically further our
understanding of the genetic inheritance of key traits in salmonids
\citep{abadia2013large,beulke2023distinct}. However, PBT is just one subset of a whole
family of statistical genetic methods employing relationship inference to learn about
populations.  For example, inference of the full siblings amongst a sample of fish
provides information about the effective number of adults producing offspring
\citep{waples2011inbreeding,wang2023estimating}, which can
be a valuable source of information when sampling of the adults is not possible.
Accordingly, marker panels should be evaluated not only on their capacity to
identify the parent-offspring relationships employed in PBT, but also on their
ability to resolve full-sibling relationships.

Here, we present a reference baseline for Chinook salmon, focusing on the regional scale of
rivers within the state of California, and particularly targeted to the complex population 
structure of Chinook salmon within California's Central Valley. Chinook salmon of the two main river basins---the 
Sacramento and the San Joaquin---within the Central Valley exhibit the greatest run-timing diversity
within the species.  With four recognized ecotypes, delineated primarily on the basis of run timing
(fall-, late-fall-, winter-, and spring-run), adult Chinook salmon can be found migrating or residing
in freshwater every month of the year in California \citep{fisher1994past}.

Chinook salmon are the largest of the Pacific salmonids and have historically been the target of extremely high-value and culturally important fisheries \citep{myers1998status}. Moreover, because of their high degree of ecotypic variation, they have provided fishery opportunities during many different seasons and geographic locations \citep{healey1991life}. However, recent declines in population sizes, from the Yukon River in Arctic north, to the southern extent of their range in California has led to almost complete fishery closures  throughout most of the northeastern Pacific to protect less productive stocks \comment{(CITE)}. The co-occurence of fish from relatively productive and relict populations reaches its paragon in California, with the largest remaining ocean fisheries for Chinook salmon targeting the Central Valley fall-run stock that spawns in different tributaries of the same river basin as the highly endangered and phenotypically distinct winter-run stock and the threatened spring-run stocks.

In spite of their distinct phenotypes and life history patterns, the four ecotypes of Chinook salmon in the Central Valley are closely related \citep{clemento2014evaluation} and share recent common ancestry that is independent of ongoing migration between these lineages. As such, GSI has been particularly challenging in the CCV, with at least one of the distinct ecotypes---late-fall-run---unresolvable with previous GSI marker sets \citep{seeb2007development,clemento2014evaluation} and with unsatisfactory power for discriminating the protected spring-run stocks from the harvested fall-run stock \comment{(CITE)}.

In the following we present a new reference baseline that includes 1636 Chinook salmon individuals
from 17 collections, genotyped at 204 amplicons from throughout the genome,
scored as microhaplotypes \citep{baetscher2018microhaplotypes}. This baseline is
highly effective for GSI within California and the markers within it provide ample power
for PBT.
We describe how we identified and compiled the markers, including several new markers
that are particularly effective at distinguishing between closely related groups of fish,
such as late-fall and fall-run Chinook salmon in the CCV.  We also provide a comprehensive analysis of this
set of markers for inference of parents and siblings within different groups of populations.
